Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-05-11 analysis/paper/paper.qmd
minor edits throughout

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages. The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, in contrast to groups with no stemmed points. Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation, such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from the Yongsandong site may indicate that the site was used for specific or limited activities, such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

@@

Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to determine their likely uses without knowing their complete shape when attached to other components. Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins]. @Lee_Kong_2002 claim stemmed points should be considered more generally as 'stemmed tools' because of the uncertainty of their complete shape. Other researchers propose that non-symmetrical stemmed points - with retouch on one side or denticulate blades on one side - should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].

@@

The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. Lombard et al. demonstrated the efficacy of TCSA by analyzing Middle Stone Age points from Sibudu Cave in KwaZulu-Natal, South Africa, which are dated to approximately 35-50 ka. They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation;@lombard2004distribution;@lombard2005evidence). @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone. We are fully acknowledged that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used. However, in the absence of detailed paleoenvironmental information, we thought it is still meaningful to adapt the modern information of vegetation and geography by assuming that some environmental factors such as  altitude, slope, drainage, hydrology, and bedrock are relatively stable over time. And those factors enable us to think that the modern vegetation patterns can be used to make reasonable references about the past.

@@

Inspired by @eren2022plains's approach, we hypothesized some simple scenarios that might explain the temporal and spatial patterns of TCSA range in Korea. If stemmed points have a narrow range of TCSA values, then people likely produced tools that performed a small set of specific tasks. This may be related to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide range of TCSA values may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable variation in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].

@@

Using weight as a size proxy, we examined the relationship between size and likely use of the stemmed points inferred from TCSA values. We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight, revealing three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50. There are a few overlaps, but we can still assume that artifacts in different size clusters might have been made for different purposes.

```{r}
#| label: fig-artifact-size-cluster
#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artifact size classes indicated by the digits representing data point values"
#| fig-width:   4
#| fig-height:  2

@@

```{r}
#| label: fig-tcsa-raw-materials
#| fig-cap: "TCSA values by lithic raw material."
#| fig-width:   4.4
#| fig-height:  3.2

@@

```{r}
#| label: fig-tcsa-radio-carbon
#| fig-cap: "A: Distribution of TCSA values over time. The light gray shade indicates Marine Isotope Stages (MIS) 3; the gray shade indicates MIS 2; and the darkest shade indicates the duration of LGM. B: Distribution of TCSA values grouped by LGM event."
#| fig-height:  5

@@

```{r}
#| label: fig-tcsa-per-sites
#| fig-cap: "TCSA values by archaeological site"
#| fig-height:  3

@@

To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TCSA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude that this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values; this was likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.

@@

```
```{r}

###### regression of CV with sample size ####

@@

tcsa_all_stage_plot <- ggplot(tcsa_all_stage, aes(x = CV, y = sample_size, color = mis_stage)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each period")+
  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
        legend.key.height = unit(0.2, 'cm'), #change legend key height
        legend.key.width = unit(0.2, 'cm'), #change legend key width
        legend.title = element_text(size=4), #change legend title font size
        legend.text = element_text(size=2)) #change legend text font size

@@

tcsa_all_raw_plot <- ggplot(tcsa_all_raw, aes(x = CV, y = sample_size, color = raw_materials_top.x)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each raw material")+
  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
        legend.key.height = unit(0.2, 'cm'), #change legend key height
        legend.key.width = unit(0.2, 'cm'), #change legend key width
        legend.title = element_text(size=4), #change legend title font size
        legend.text = element_text(size=2)) #change legend text font size

@@

tcsa_all_ass_plot <- ggplot(tcsa_all_ass, aes(x = CV, y = sample_size, color = Assemblage.x)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each asseblage(Location)") +
  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
        legend.key.height = unit(0.2, 'cm'), #change legend key height
        legend.key.width = unit(0.2, 'cm'), #change legend key width
        legend.title = element_text(size=4), #change legend title font size
        legend.text = element_text(size=2)) #change legend text font size

@@

tcsa_all_veg_plot <- ggplot(tcsa_all_veg, aes(x = CV, y = sample_size, color = ecoregion)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each vegetation zone")+
  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
        legend.key.height = unit(0.2, 'cm'), #change legend key height
        legend.key.width = unit(0.2, 'cm'), #change legend key width
        legend.title = element_text(size=4), #change legend title font size
        legend.text = element_text(size=2)) #change legend text font size

@@

tcsa_all_geo_plot <- ggplot(tcsa_all_geo, aes(x = CV, y = sample_size, color = Zone)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each geographic zone")+
  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
        legend.key.height = unit(0.2, 'cm'), #change legend key height
        legend.key.width = unit(0.2, 'cm'), #change legend key width
        legend.title = element_text(size=4), #change legend title font size
        legend.text = element_text(size=2)) #change legend text font size

plot_grid(  tcsa_all_stage_plot,
  tcsa_all_raw_plot,
tcsa_all_ass_plot,
 tcsa_all_veg_plot, 
  tcsa_all_geo_plot,
ncol =2,
label_size = 1)

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that fall within the ranges of dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to the unavailability of their data. They claimed that these values are within the range of North American dart tips and arrowheads. We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads. In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explored the relationship between TCSA for weight and raw material types in order to examine these factors. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). There was no clear pattern for raw materials (@fig-tcsa-raw-materials). However, combined with weight, we observed a positive relationship between raw material type and TCSA values (@fig-tcsa-size). We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels within the landscape. Therefore, we find that raw materials, via nodule size, were influential on TCSA values.

@@

We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) Zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as the CT Zone, show higher variability. These results suggest that stemmed points performed a wider range of tasks in low-productivity patches. @prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points. We similarly found that stemmed points are more densely distributed in certain environments in South Korea. Future work should investigate more specifically patterns in the distribution of stemmed points and ranges of faunal taxa.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages excavated from `r n_sites` sites. Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points likely reflects the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were newly introduced [@chang2013human], in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.

We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we plan to conduct use-wear analyses and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.

>>>>

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages. They found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, in contrast to groups with no stemmed points. Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation, such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from the Yongsandong site may indicate that the site was used for specific or limited activities, such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

@@

Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it has been difficult to determine their likely uses without knowing their complete shape when attached to other components. Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins]. @Lee_Kong_2002 claim stemmed points should be considered more generally as 'stemmed tools' because of the uncertainty of their complete shape. Other researchers propose that non-symmetrical stemmed points - with retouch on one side or denticulate blades on one side - should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].

@@

The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. Lombard et al. demonstrated the efficacy of TCSA by analyzing Middle Stone Age points from Sibudu Cave in KwaZulu-Natal, South Africa, which are dated to approximately 35-50 ka. They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation; @lombard2004distribution; @lombard2005evidence). @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone. We note that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used. Unfortunately, reconstructions of Pleistocene landscapes are not available for our study area. We assume the underlying environmental factors such as altitude, slope, drainage, hydrology, and bedrock are relatively stable over time, and are suitable as an analogy for initial hypothesizing about relationships between Pleistocene geographical variation and stone artefact technology.

@@

Inspired by @eren2022plains's approach, we hypothesized simple scenarios that might explain the temporal and spatial patterns of TCSA range in Korea. If stemmed points have a narrow range of TCSA values, then people likely produced tools that performed a small set of specific tasks. This may be related to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide range of TCSA values may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable variation in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].

@@

Using weight as a size proxy, we examined the relationship between size and likely use of the stemmed points inferred from TCSA values. We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight, revealing three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.

```{r}
#| label: fig-artifact-size-cluster
#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artifact size classes indicated by the digits representing data point values. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1"
#| fig-width:   4
#| fig-height:  2

@@

```{r}
#| label: fig-tcsa-raw-materials
#| fig-cap: "TCSA values by lithic raw material. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
#| fig-width:   4.4
#| fig-height:  3.2

@@

```{r}
#| label: fig-tcsa-radio-carbon
#| fig-cap: "A: Distribution of TCSA values over time. The light gray shade indicates Marine Isotope Stages (MIS) 3; the gray shade indicates MIS 2; and the darkest shade indicates the duration of LGM. B: Distribution of TCSA values grouped by LGM event. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
#| fig-height:  5

@@

```{r}
#| label: fig-tcsa-per-sites
#| fig-cap: "TCSA values by archaeological site. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
#| fig-height:  3

@@

To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map and Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed some statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TCSA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude that this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values; this was likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.

@@

```{r}
#| eval: false

###### regression of CV with sample size ####
# To address the question that CV is controlled more by sample size than anything else, 
# this does not produce a figure in the paper currently, just for our satisfaction

@@

@@

@@

@@

@@

# BM combine data frames so we can put them all on a single plot panel

library(ggrepel)
bind_rows(tcsa_all_stage %>% rename(var = mis_stage),
          tcsa_all_raw   %>% rename(var = raw_materials_top.x),
          tcsa_all_ass   %>% rename(var = Assemblage.x ),
          tcsa_all_veg   %>% rename(var = ecoregion ),
          tcsa_all_geo   %>% rename(var = Zone )) %>% 
  ggplot() + 
         aes(x = sample_size, 
             y = CV, 
             label = var) +
  geom_point() +
  geom_text_repel() +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(y = "CV of Min TCSA", 
       x = "Sample size") +
  theme_minimal()

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that fall within the ranges of dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to the unavailability of their data. They claimed that these values are within the range of North American dart tips and arrowheads. We found a wider range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads. In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that lower TCSA values could be the result of pursuing production economy and transport efficiency. We explored the relationship between TCSA for weight and raw material types in order to examine these factors. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). There was no clear pattern for raw materials (@fig-tcsa-raw-materials). However, combined with weight, we observed a positive relationship between raw material type and TCSA values (@fig-tcsa-size). We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels within the landscape. Therefore, we find that raw materials, via nodule size, were influential on TCSA values.

@@

We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) Zone with few clear patterns in TCSA function evident across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as the CT Zone, show higher variability. These results suggest that stemmed points performed a wider range of tasks in low-productivity patches. @prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points. We similarly found that stemmed points are more densely distributed in certain environments in South Korea. Future work should investigate more specifically patterns in the distribution of stemmed points and ranges of faunal taxa.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What were the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages excavated from `r n_sites` sites. Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the uses of stemmed points likely reflects the socio-environmental circumstances that people encountered. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were introduced [@chang2013human] in response to global climate dynamics, and stemmed points were optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.

We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation]. Use-wear analyses and experiments should be part of future research into the function of stemmed points to investigate the validity of our current results.
